---
title: "Automobile Gasoline Mileage Analysis"
author: "Yasser Gonzalez --- http://yassergonzalez.com"
date: "July 2015"
graphics: yes
output:
  pdf_document: default
---

# Introduction

For additional information, see Henderson and Velleman [1].

# Exploratory Data Analysis

```{r packages, echo = FALSE, results = "hide"}
library("knitr")
library("dplyr", warn.conflicts = FALSE)
library("ggplot2")
```

The data set of interest was loaded into R and tidied (see the R code in the 
Appendix). The following table shows six selected rows from the resulting 
data set.

```{r dataset, echo = FALSE}
data("mtcars")

mtcars <- mtcars %>%
    add_rownames("car") %>%
    mutate(am = ifelse(am == 0, "automatic", "manual"))

kable(head(mtcars, n = 6))
```

Given that we are interested in comparing gas mileage of cars with automatic 
or manual transmission, the analysis begins with a boxplot of the 
associated mpg distributions.

```{r boxplot, fig.width = 5, fig.height = 2.5, fig.align = "center", echo = FALSE}
ggplot(mtcars, aes(x = am, y = mpg)) +
    geom_boxplot() +
    labs(x = NULL) +
    coord_flip()
```

The plot suggests that, among the group of studied cars, manual transmission
cars have better gas mileage (as in higher mpg values) than cars with 
automatic transmission. This claim is studied in more detail in the next
section with a regression analysis.

# Regression Models Analysis

# Conclusions

# References

1.  Henderson and Velleman (1981). Building multiple regression models 
    interactively. *Biometrics*, **37**, pages 391-411.


# Appendix

This appendix shows the R code fragments executed to obtain the 
reported results.

1. Loading the necessary R packages.

```{r ref.label = "packages", eval = FALSE}
```

2. Loading and tidying the data set.

```{r ref.label = "dataset", eval = FALSE}
```

3. Boxplot of mpg.

```{r ref.label = "boxplot", eval = FALSE}
```
