---
title: "Automobile Gasoline Mileage Analysis"
author: "Yasser Gonzalez --- http://yassergonzalez.com"
date: "July 2015"
graphics: yes
output:
  pdf_document: default
---

# Introduction

`mtcars` data set.

See the following publication for additional information about this data set:

> Henderson and Velleman (1981). "Building multiple regression models 
> interactively". *Biometrics*, **37**, pages 391-411.

# Exploratory Data Analysis

```{r packages, echo = FALSE, results = "hide"}
library("knitr")
library("dplyr", warn.conflicts = FALSE)
library("ggplot2")
```

The data set of interest was loaded into R and tidied (see the R code in 
the Appendix). The following table shows six selected rows from the 
resulting data set.

```{r dataset, echo = FALSE}
library("datasets")
data("mtcars")

mtcars <- mtcars %>%
    add_rownames("car") %>%
    mutate(am = ifelse(am == 0, "Automatic", "Manual")) %>%
    mutate(vs = ifelse(vs == 0, "V-engine", "Straight engine"))
```

```{r echo = FALSE}
kable(head(mtcars, n = 6))
```

Given that we are interested in comparing gas mileage of cars with automatic 
or manual transmission, we begin the analysis with a boxplot of the 
associated `mpg` distributions.

```{r boxplot, fig.width = 5, fig.height = 2.5, fig.align = "center", echo = FALSE}
ggplot(mtcars, aes(x = am, y = mpg)) +
    labs(x = NULL, y = "Miles per gallon") +
    geom_boxplot() +
    coord_flip()
```

The plot suggests that, among the group of studied cars, manual transmission
cars have better fuel efficiency (as in higher `mpg` values) than cars with 
automatic transmission. This claim is studied in more detail with a regression 
analysis in the next section.

# Regression Models Analysis

Motivated by what was observed in the boxplot, we construct first a linear model for 
explaining the `mpg` values in terms of the `am` variable (i.e. the transmission type) 
coded as a dummy variable as follows:

```{r only_am}
only_am <- lm(mpg ~ am, data = mtcars)
```

```{r include = FALSE}
only_am_coef <- summary(only_am)$coefficients
```

The fitted coefficients table is shown in the Appendix. The coefficient 
corresponding to the `am` dummy variable associated with manual transmission 
in effect indicates that manual transmission cars have better fuel efficiency 
than automatic cars---roughly `r round(only_am_coef[2, 1], 1)` miles per gallon 
better on average. The associated $p$-value also suggests that this result is 
statistically significant at the 1% level. However, the $R^2$ obtained is 
`r round(summary(only_am)$r.squared, 2)`, indicating that the model leaves
most of the variance of `mpg` unexplained. This is not a surprising result, 
considering that this model predicts the same `mpg` value for all manual 
transmission cars; and similarly, the same `mpg` value for all cars 
with automatic transmission. These results suggest that the model denoted as 
`only_am` is a poor fit for explaining the cars' fuel efficiency.

We construct another linear model next. It considers all the other variables 
in the data set as predictors (in addition to `am`) in order to understand 
if there are other factors that influence the `mpg` values. The second linear 
model can be fitted as follows:

```{r all_vars}
all_vars <- lm(mpg ~ am + ., data = select(mtcars, -car))
```

```{r include = FALSE}
all_vars_coef <- summary(all_vars)$coefficients
```

In this case, the obtained $R^2$ score is approximately 
`r round(summary(all_vars)$r.squared, 2)`, which means that the model explains 
roughly `r round(100 * summary(all_vars)$r.squared, 0)`% of the variance of 
the `mpg` values. A likelihood ratio tests for nested 
models comparing `only_am` and `all_vars` indicates a statistically
significant reduction at the 1\% level of the residual sum of squares
by including the other variables (see the details in the Appendix). 
Also, a plot included in the Appendix illustrates an appropriate 
distribution of the residuals. 

These results evidence that the `all_vars` linear model is a better fit 
to explain the `mpg` values than the `only_am` model---i.e. that other 
variables influence the fuel efficiency of the cars besides the transmission 
type. In fact, the coefficient corresponding to the `am` dummy variable 
associated with manual transmission in the `all_vars` model is approximately 
`r round(all_vars_coef[2, 1], 2)` and it is not statistically significantly
different from zero at the 1\% level. 

Summarizing, the transmission type does not seem to have a significant impact 
on the fuel efficiency of the cars after having accounted for the effect of 
other variables such as the number of cylinders and the weight of the car.

# Appendix

This appendix contains supplementary figures and the R code fragments not shown 
in the main section to ease the reproducibility of the results.

1. Loading the necessary R packages.

```{r ref.label = "packages", eval = FALSE}
```

2. Loading and tidying the data set.

```{r ref.label = "dataset", eval = FALSE}
```

3. Code to generate the `mpg` boxplot.

```{r ref.label = "boxplot", eval = FALSE}
```

4. Coefficients table for the `only_am` model.

```{r echo = FALSE}
kable(only_am_coef)
```

5. Coefficients table for the `all_vars` model.

```{r echo = FALSE}
kable(all_vars_coef)
```

6. Comparison of the `only_am` and `all_vars` linear models.

```{r eval = FALSE}
anova(only_am, all_vars)
```

```{r echo = FALSE}
kable(anova(only_am, all_vars))
```

7. Plot of the residuals of the `all_vars` linear model.

```{r fig.width = 5, fig.height = 3, fig.align = "center"}
data <- data.frame(residuals = residuals(all_vars),
                   fitted_values = predict(all_vars))
ggplot(data, aes(x = fitted_values, y = residuals)) +
    labs(x = "Fitted values",  y = "Residuals") +
    geom_point()
```
